---
title: "Presentation"
author: "Asbjørn, Emiliano, Magnus, Oliver and Trine"
format:
  revealjs:
    theme: simple
    transition: slide
    footer: "Group 01 R for Bio data science 22160"
    highlight-style: github
    code-fold: false
output-dir: doc
---

# Introduction 2

-   Dataset from NCBI
-   Transcriptional profiling of colon epithelial from ulcerative colitis patients and healthy controls
-   Ulcers in the colon
    -   chronic inflamation
-   Biopsies from the gut
    -   202 samples
    -   from 67 patients
    -   30 healthy controls
-   Gene expression
-   Aim to find and analyze variation between healthy controls and patients
    -   what smoking does
    -   Effect of genes

# Materials of Methods 3

::::::: columns
::: {#Plot 2 .column width="40%"}
Maybe put plot of age distribution / histogram
:::

::::: columns
::: {.column width="40%"}
:::

::: {.column width="20%"}
Noble CL, Abbas AR, Cornelius J, et al

Regional variation in gene expression in the healthy colon is dysregulated in ulcerative colitis

Gut 2008;**57:**1398-1405.
:::
:::::
:::::::

## Filtering data 4

::::: columns
::: {.column width="50%"}
Removing one-type columns

```{r Cleaning_phenodata, echo=TRUE, eval=FALSE}
phenodata <- phenodata |>
  select(
    where(
      ~ n_distinct(
        .x, na.rm = TRUE) > 1))
```
:::

::: {.column width="50%"}
Removing low variance genes

```{r Cleaning_genes, echo=TRUE, eval=FALSE}
gene_var <- exprs_df |> 
  # Drop the sample ID column, keep only expression values
  select(-geo_accession) |> 
  summarise(across(everything(), ~ var(.x, na.rm = TRUE))) |> 
  pivot_longer(
    cols      = everything(),
    names_to  = "gene",
    values_to = "variance"
  )
# Cutoff genes that does not vary more than 0.05 - 2909 genes remaining
genes_keep <- gene_var |> 
  filter(variance > 0.05) |> 
  pull(gene)

exprs_filtered <- exprs_df |> 
  select(geo_accession, all_of(genes_keep))

```
:::
:::::

## Joining Data 5 

## Augmenting Data 6

# Results 7 

# plot 8

# Discussion 9

## Correlations and Causality 10

