---
title: "08_analysis_umap"
format: html
---

# Analysis 4

## Loading libraries

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(randomForest)
set.seed(6969)
```

```{r}
#| message: false
#| warning: false

expr_df <- read_csv("../data/03_rf_expr_df.csv.gz")
featuredata <- read_csv("../data/03_dat_genbank.csv.gz")
de_results <- read_csv("../data/03_de_results.csv.gz")
```

## Model Training

### Splitting dataset into train and test

```{r}
percent_train <- 0.8 #use 80% of the data for train 
train_index <- sample(1:nrow(expr_df), 0.8 * nrow(expr_df))

train_data <- expr_df[train_index, ]
test_data <- expr_df[-train_index, ]
```

### Running the RF Classifier model

```{r}
#y_pred_UC as factor was not implemented in 03_augment, despite the use of as.factor()
train_data <- train_data |> 
  mutate(y_pred_UC = as.factor(y_pred_UC))
```

```{r}
# y_pred_UC ~ . 
train_data.rf <- randomForest(y_pred_UC ~ . - geo_accession, 
                              type = "classification",
                              data = train_data, 
                              mtry = 3,
                              importance = TRUE)
```

## Model Evaluation

### Deriving the Confusion Matrix of the RF model

```{r}
test_pred <- predict(train_data.rf, test_data)
confusion_matrix <- table(Predicted = test_pred, Actual = test_data$y_pred_UC)
print(confusion_matrix)
```

Confusion matrix shows the amount of times the model predicted 0 or 1 (rows), based on the actual values of 0 or 1 in the test dataset. It therefore shows:

|     |     |
|:---:|:---:|
| TN  | FN  |
| FP  | TP  |

where;

-   TN: True Negative

-   FN: False Negative

-   FP: False Positive

-   TP: True Positive

From the matrix, we can obtain an accuracy score.

```{r}
#Correct predictions; TN, TP
TN <- confusion_matrix[1, 1]
TP <- confusion_matrix[2, 2]
FP <- confusion_matrix[2, 1]
FN <- confusion_matrix[1, 2]

total <- sum(confusion_matrix)
acc <- (TN+TP)/total
print(acc)
```

The model emitted a 78% accuracy.

### Importance measure of Genes

```{r}
var_importance <- as.data.frame(round(importance(train_data.rf), 2))
```

```{r}
var_importance <- var_importance |> 
  mutate(gene_id = str_split_i(rownames(var_importance), "_", 2))
```

```{r}
var_importance |> 
  #mutate(gene_id = str_split_i(rownames(var_importance), "_", 2)) |> 
  arrange(desc(MeanDecreaseAccuracy)) |> 
  slice(1:15) |>
  ggplot(aes(
    # 1. Use reorder() to automatically sort the bars by the MeanDecreaseAccuracy value
    #x = rownames(var_importance),
    x = reorder(gene_id, MeanDecreaseAccuracy), 
    y = MeanDecreaseAccuracy
  )) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() + 
  labs(
    title = "Top 15 Most Contributing Genes in RF",
    x = "Gene",
    y = "Mean decrease accuracy from RF"
  ) +
  theme_minimal()
```

Extract gene_ids as list

```{r}
most_contributing_vars <- var_importance |> 
  arrange(desc(MeanDecreaseAccuracy)) |> 
  slice(1:15) |> 
  rownames()
```

### Match gene_id with gene_symbol to match other analysis

GENE_SYMBOL can be found in featuredata

```{r}
#Splitting gene_id into only the number and converting to numeric
most_contributing_vars <- most_contributing_vars |> 
  str_split_i("_", 2) |> 
  parse_number()

most_contributing_vars <- as.data.frame(most_contributing_vars)
most_contributing_vars <- most_contributing_vars |> 
  mutate(ID = most_contributing_vars) |> 
  select(ID)
```

```{r}
most_contributing_gene_symbols <- featuredata |>
  select(ID, GENE_SYMBOL) |> 
  right_join(most_contributing_vars, featuredata, by="ID") |> 
  select(GENE_SYMBOL) |>
  filter(!is.na(GENE_SYMBOL)) |> 
  pull(GENE_SYMBOL)

print(most_contributing_gene_symbols)
```

## Redo vulcano plot from 02

To visualize overlay between significant genes found in 05_analysis, and the RandomForest model most contributing variables.

The 05_analysis relied on inflammation status, and the model relied on disease state.

```{r}
#| warning: false
#| message: false
de_results <- read_csv("../data/03_de_results.csv.gz")
```

```{r}
lfc_cut <- 0.25   # minimum |log2FC| to call biologically relevant
p_cut   <- 0.05   # p-value threshold for the dashed line

## 1) Add gene symbol by matching finaldata$gene_id --> featuredata$ID
de_results_annot <- de_results |>
  mutate(
    GENE_SYMBOL = featuredata$GENE_SYMBOL[
      match(gene_id, featuredata$ID)
    ]
  )

## 2) Label only highly significant genes with valid symbols
label_df <- de_results_annot |>
  filter(
    neg_log10_p > 10,          # very strong p-values
    !is.na(GENE_SYMBOL)
  )

## 3) Prepare plotting data: FDR + 3-level significance category
plot_df <- de_results_annot |>
  filter(!is.na(p_value)) |>
  mutate(
    FDR = p.adjust(p_value, method = "BH"),
    sig_cat = case_when(
      FDR < 0.05 & log2FC >=  lfc_cut ~ "Up in inflamed",
      FDR < 0.05 & log2FC <= -lfc_cut ~ "Up in uninflamed",
      TRUE                           ~ "Non significant"
    ),
    sig_cat = factor(sig_cat,
                     levels = c("Up in inflamed", "Up in uninflamed", "Non significant")),
    
    # Identify genes from the highlight list
    highlight_cat = case_when(
      GENE_SYMBOL %in% most_contributing_gene_symbols ~ "Most Contributing",
      TRUE ~ "Other"
    )
  )

sig_genes <- plot_df |>
  filter(sig_cat != "Non significant") |>
  dplyr::select(gene_id, GENE_SYMBOL, log2FC, p_value, FDR, sig_cat) |>
  arrange(FDR)

# split into up in inflamed / up in uninflamed
up_in_inflamed <- sig_genes |>
  filter(sig_cat == "Up in inflamed") |>
  pull(GENE_SYMBOL)

up_in_uninflamed <- sig_genes |>
  filter(sig_cat == "Up in uninflamed") |>
  pull(GENE_SYMBOL)

## 4) Volcano plot with Highlighted Genes
ggplot(plot_df,
       aes(x = log2FC, y = neg_log10_p)) +

  # 1. Base points (coloured by sig_cat)
  geom_point(aes(color = sig_cat), size = 1.3, alpha = 0.8) +

  # 2. Overlay highlighted genes (colored by highlight_cat)
  geom_point(data = filter(plot_df, highlight_cat == "Most Contributing"),
             aes(color = highlight_cat),
             size = 1.3) +

  # cutoff lines
  geom_hline(yintercept = -log10(p_cut),
             linetype = "dashed", linewidth = 0.3) +
  geom_vline(xintercept = c(-lfc_cut, lfc_cut),
             linetype = "dashed", linewidth = 0.3) +

  # labels for strongest hits
  geom_text_repel(
    data = label_df,
    aes(label = GENE_SYMBOL),
    size = 3,
    colour = "black",
    max.overlaps = 30,
    box.padding = 0.4,
    point.padding = 0.3,
    segment.size = 0.2,
    show.legend = FALSE
  ) +

  # unified color scale (sig categories + highlighted category)
  scale_color_manual(
    values = c(
      "Up in inflamed"   = "#d73027",
      "Up in uninflamed" = "#4575b4",
      "Non significant"  = "grey80",
      "Most Contributing" = "green"
    ),
    name = NULL,
    labels = c(
      "Up in inflamed"   = "Up in inflamed",
      "Up in uninflamed" = "Up in uninflamed",
      "Non significant"  = "Non significant",
      "Most Contributing" = "Genes present in RF model"
    )
  ) +

  labs(
    title = "Volcano plot of colonic gene expression",
    subtitle = "Inflamed vs uninflamed biopsies (Gut 57:1398–1405)",
    x = "log2 fold change (Inflamed – Uninflamed)",
    y = expression(-log[10](italic(p)~value))
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(size = 12)
  )
```

## Cleaning environment

```{r}
rm(list = ls())
```
